---
title: "ICE activity and Hispanic school enrollment in MD Memo - Will Hammann"
author: "Hammann"
date: "2026-06-01"
output: html_document
---
##INTRO

# load libraries
```{r}
# install.packages("tidyverse")
# install.packages("tidycensus")
# install.packages("janitor")
library(tidyverse)
library(tidycensus)
library(janitor)
```

#load and clean data (and combining)
```{r}
# There are files with the same naming format for 2021-2026 in the same location. Only files for the years 2020-2022 require renaming lss to lea and lss_name to lea_name
# After cleaning datasets of all years, they can be easily combined and cleaned a bit more so columns are strings or numeric where they should be.
data_2020<-read_csv("../data/Enrollment_By_Race_2020.csv") |>
     clean_names() |> 
     rename(lea = lss, lea_name = lss_name)

data_2021<-read_csv("../data/Enrollment_By_Race_2021.csv") |>
     clean_names() |> 
     rename(lea = lss, lea_name = lss_name)

data_2022<-read_csv("../data/Enrollment_By_Race_2022.csv") |>
     clean_names() |>
     rename(lea = lss, lea_name = lss_name)

data_2023<-read_csv("../data/Enrollment_By_Race_2023.csv") |>
     clean_names()

data_2024<-read_csv("../data/Enrollment_By_Race_2024.csv") |>
     clean_names()

data_2025<-read_csv("../data/Enrollment_By_Race_2025.csv") |>
     clean_names()

data_2026<-read_csv("../data/Enrollment_By_Race_2026.csv") |>
     clean_names()

data_combined<-bind_rows(list(data_2020, data_2021, data_2022, data_2023, data_2024, data_2025, data_2026))

data_combined$enrolled_count[data_combined$enrolled_count == '*'] <- NA


data_combined<-data_combined |>
  mutate(
    enrolled_count = as.numeric(enrolled_count),
    across(where(is.character), str_to_lower),
    create_date = as.Date(as.character(create_date), format = "%Y%m%d"),
    lea_name = str_replace_all(lea_name, "'", ""),
    school_name = str_replace_all(school_name, "'", "")
  )
```

**FINDING** All counties but XX saw significant growth in Hispanic student enrollment in 2024. 

Maryland and DMV outlets reported a drop of 4.4% in the graduation rate of Hispanic students in the state according to School Board President Josh Michael. 

**NOTE** 
Data is created near the start of the calendar year it represents. E.g., percent change for 2026 is the change from the beginning of 2025 to beginning of 2026. 


```{r}
options(scipen = 999)  # Disable scientific notation globally

state_data <- data_combined |> 
  filter(
    lea_name == "state",
    race == "hispanic/latino of any race",
  ) |>
  group_by(year) |> 
    summarize(
      attendance = sum(enrolled_count, na.rm = TRUE),
      .groups = "drop"
  ) |> 
  arrange(year) |>
  mutate(
    pct_change = 100 * (attendance / lag(attendance) - 1)
  )

counties_data <- data_combined |> 
  filter(
    race == "hispanic/latino of any race",
    school != "a"
  ) |>
  group_by(lea_name, year) |>
  summarize(
    attendance = sum(enrolled_count, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  arrange(lea_name, year) |>
  mutate(
    pct_change = 100 * (attendance / lag(attendance) - 1)
  )

state_data |>
  filter(year != "2020")

counties_data |>
  filter(year != "2020")
```


## Statistical research by Stanford professors showed a 7.3% decrease in Hispanic enrollment in counties ICE partnerships with local police nationwide in the 2000s, establishing precedent for...

## Including but not exclusive to his administration, I can also look at the correlation between Hispanic enrollment and 287(g) agreements in counties or even municipalities.
```{r}
dhs_287gs<-read.csv("../data/287g_agreements_dhs.csv") |>
  clean_names()
```


## Potential Sources
# 


## Attendance by Race by School for a Specific County in 2020
```{r}
data_2020 |> 
  filter(lea_name == "Calvert") |> 
  group_by(school_name, race) |>
  summarize(
    Attendance = sum(enrolled_count, na.rm = TRUE)
  ) |> 
  arrange(school_name, race)

```
