---
title: "ICE activity and Hispanic school enrollment in MD Memo - Will Hammann"
author: "Hammann"
date: "2026-06-01"
output: html_document
---

##The data shows a XX enrollment in Prince George's and Montgomery counties from YYYY to YYYY, the two highest concentrations of Hispanic residents among MD counties. 

## There was a 4.4% drop in graduation rates for Hispanic students and a 5.5% drop for what education officials categorize as multilanguage learners in MD schools in 2025. VERIFY WITH DATA

##Statistical research by Stanford professors showed a 7.3% decrease in Hispanic enrollment in counties ICE partnerships with local police nationwide in the 2000s, establishing precedent for...

I plan to look at the level of Hispanic student enrollment in Maryland schools with high Hispanic populations post-immigration crackdown of the Trump administration. Including but not exclusive to his administration, I can also look at the correlation between Hispanic enrollment and 287(g) agreements in counties or even municipalities. The data is from 2020-26 and includes individual schools so I can get deep into standout examples after further analysis. 

## Potential Sources
# 


# load libraries
```{r}
# install.packages("tidyverse")
# install.packages("tidycensus")
# install.packages("janitor")
library(tidyverse)
library(tidycensus)
library(janitor)

```

#load and clean data
```{r}
# there are files with the same naming format for 2021-2026 in the same location. Only files for the years 
# 2020-2022 require renaming lss to lea and lss_name to lea_name
data_2020<-read_csv("../data/Enrollment_By_Race_2020.csv") |>
     clean_names() |> 
     rename(lea = lss, lea_name = lss_name)

data_2020$enrolled_count[data_2020$enrolled_count == '*'] <- NA
data_2020$enrolled_count <- as.numeric(data_2020$enrolled_count)

```

## Attendance by Race for 2020

```{r}
options(scipen = 999)  # Disable scientific notation globally

data_2020 |> 
group_by(race) |> 
  summarize(
    Attendance = sum(enrolled_count, na.rm = TRUE)
  ) |> 
  arrange(race)

```

## Attendance by Race by County for 2020

```{r}
options(scipen = 999)  # Disable scientific notation globally

data_2020 |> 
group_by(lea_name, race) |> 
  summarize(
    Attendance = sum(enrolled_count, na.rm = TRUE)
  ) |> 
  arrange(lea_name, race)

```

## Attendance by Race by School for a Specific County in 2020
```{r}
data_2020 |> 
  filter(lea_name == "Calvert") |> 
  group_by(school_name, race) |>
  summarize(
    Attendance = sum(enrolled_count, na.rm = TRUE)
  ) |> 
  arrange(school_name, race)

```

```{r}

```

